Lexical frequency mediates the facilitation effect of cognateness in bilingual lexical acquisition

### 2023-04-28 _appendix/_model-details.qmd
Add appendix for syllable frequencies and new revisions

where:

- $y$ is an observed response ($y \in \{\text{No, Understands, Understands and Says}\}$)
- $i$ is the participant index
- $j$ is the translation equivalent (TE) index
- $p_{k}$ is a probability ($p \in (0, 1)$) that indicates the threshold $k$ ($k \in (1, 2)$)) between two response categories in the latent distribution
$p_{k}$ is then estimated using a logit regression model as indicated in @eq-linear.

To test our hypotheses, we included several predictors in the regression model as fixed effects: the main effects of participant age ($\text{Age}$), word-form length ($\text{Length}$), word exposure index ($\text{Exposure}$), and of phonological similarity ($\text{Cognateness}$), the two-way interactions $\text{Age} \times \text{Exposure}$, $\text{Age} \times \text{Cognateness}$, and $\text{Exposure} \times text{Leveshtein}$, and the three-way interaction $\text{Age} \times \text{Exposure} \times \text{Leveshtein}$. We also included crossed random effects for participants and translation equivalents to account for the repeated measures in our dataset---each participant provided responses to multiple translation equivalents, and each translation equivalent was responded to by multiple participants [@gelman2020regression]. For both grouping variables, we included random intercepts, random slopes, and correlation parameters for all predictors were repeated measures were observed in our dataset [@barr2013random, see @eq-linear].

$$
\begin{aligned}
\textbf{Linear model:} \\
logit(p_{k}) = \text{ln} \frac{p_{k}}{1-p_{k}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + \\
& (\beta_{1} + u_{1_{i}} + w_{1_{j}}) · \text{Age}_{i} + & \\
& (\beta_{2} + u_{2_{i}} + w_{2_{j}}) · \text{Length}_{ij} + & \\
& (\beta_{3} + u_{3_{i}} + w_{3_{j}}) · \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) · \text{Cognateness}_{ij} + & \\
& (\beta_{5} + u_{5_{i}} + w_{3_{j}}) · (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) · (\text{Age}_{i} \times \text{Cognateness}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) · (\text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) · (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
\end{aligned}
$$ {#eq-linear}

where:

- $i$ and $j$ index the participant and translation equivalent (TE)
- $\beta_{0_k}$ is the fixed coefficient of the regression model for the intercept of threshold $k$
- $u_{0_{i}}$ and $w_{0_{j}}$ are the by-participant and by-TE adjustments for $\beta_{0_{k}}$ (i.e., random intercepts), respectively
- $\beta_{1-8}$ are the fixed coefficients of the regression model for the predictors of interest
- $u_{1-8_{i}}$ and $w_{1-3_{j}}$ are the by-participant and by-TE adjustments for$\beta_{1-8}$ (i.e., random slopes), respectively

We used the Bayesian framework to estimate the parameters in our model. This involves using the Bayes theorem to compute a distribution (*posterior distribution*) that describes what values of each parameter in the model are more likely given the data (*likelihood*, see @eq-likelihood), and previous knowledge about such distribution (*prior distribution*, see @eq-prior) [@mcelreath2020statistical]. This posterior distribution not only informs about the most likely values of our regression coefficients of interest, but also about the uncertainty around such estimations. We used a weakly informative prior for our parameters, with the exception of the main effect of $\text{Age}$, for which we specified a strongly informative prior based on previous literature about how age affects the acquisition of words [see @eq-prior].

@@

where:

- $\rho_{u_{0-8}}$ and $\rho_{w_{0-3}}$ indicate the correlation parameters between the by-participant and by-TE adjustments, respectively
- $\sigma_{u_{0-8}}^2$ and $\sigma_{w_{0-3}}^2$ indicate the variance of the by-participant and by-TE variance of the adjustments, respectively
- $\mathcal{N}$ indicates a normal distribution, $\mathcal{N}_{+}$ indicates a truncated normal distribution with only positive values, and $LKJcorr$ indicates a [LKJ correlation distribution](https://mc-stan.org/docs/2_22/functions-reference/lkj-correlation.html) [@lewandowski2009generating].

```{r}
#| label: fig-prior-mean
#| echo: false
#| message: false
#| warning: false
#| eval: false
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Expected prior predictions."
nd <- expand.grid(
    n_phon_std = 0,
    age_std = (10:36-mean(responses$age))/(sd(responses$age)),
    exposure_std = 0,
    lv_std = 0
)

m <- add_epred_draws(nd, 
                     model_fit_4_prior,
                     ndraws = NULL, 
                     re_formula = NA) |> 
    filter(.category != "No") |> 
    pivot_wider(names_from = ".category", values_from = ".epred") |> 
    mutate(Understands = Understands + `Understands and Says`) |> 
    pivot_longer(
        c(Understands, `Understands and Says`),
        names_to = ".category",
        values_to = ".epred"
    ) |> 
    mutate(
        .category = factor(
            .category,
            levels = c("Understands and Says", "Understands"),
            labels = c("Production\n(Understands and Says)", "Comprehension\n(Understands)")
        )
    )

m |> 
    ggplot(aes(x = age_std, y = .epred)) +
    facet_wrap(~.category) +
    stat_lineribbon(linewidth = 1, 
                    colour = NA,
                    .width = c(0.95, 0.89, 0.67, 0.5)) +
    stat_summary(fun = "mean", geom = "line", linewidth = 0.75) +
    scale_x_continuous(
        breaks = scale(seq(10, 36, 4), 
                       mean(responses$age), 
                       sd(responses$age))[, 1],
        labels = seq(10, 36, 4)
    ) +
    scale_fill_manual(
        values = rev(clrs), 
        labels = c("95%", "89%", "67%", "50%")
    ) +
    labs(
        x = "Age (months)", 
        y = "Probability of acquisition",
        colour = "Credible interval", 
        fill = "Credible interval"
    ) +
    scale_y_continuous(labels = percent, limits = c(0, 1)) +
    theme(
        legend.position = "top",
        axis.ticks.x = element_line(colour = "black"),
        panel.grid = element_blank()
    )
```

>>>>

where: $y$ is an observed response ($y \in \{\text{No, Understands, Understands and Says}\}$), $i$ is the participant index, $j$ is the translation equivalent (TE) index, $p_{k}$ is a probability ($p \in (0, 1)$) that indicates the threshold $k$ ($k \in (1, 2)$) between two response categories in the latent distribution $p_{k}$ is then estimated using a logit regression model as indicated in @eq-linear.

To test our hypotheses, we included several predictors in the regression model as fixed effects: the main effects of participant age ($\text{Age}$), word-form length ($\text{Length}$), word exposure index ($\text{Exposure}$), and of phonological similarity ($\text{Cognateness}$), the two-way interactions $\text{Age} \times \text{Exposure}$, $\text{Age} \times \text{Cognateness}$, and $\text{Exposure} \times \text{Cognateness}$, and the three-way interaction $\text{Age} \times \text{Exposure} \times \text{Cognateness}$. We also included crossed random effects for participants and translation equivalents to account for the repeated measures in our dataset---each participant provided responses to multiple translation equivalents, and each translation equivalent was responded to by multiple participants [@gelman2020regression]. For both grouping variables, we included random intercepts, random slopes, and correlation parameters for all predictors were repeated measures were observed in our dataset [@barr2013random], see (@eq-linear).

$$
\begin{aligned}
\textbf{Linear model:} \\
\text{ln} \frac{p_{k}}{1-p_{k}} &= (\beta_{0_{k}} + u_{0_{i_{k}}} + w_{0_{j_{k}}}) + \\
& (\beta_{1} + u_{1_{i}} + w_{1_{j}}) \cdot \text{Age}_{i} + & \\
& (\beta_{2} + u_{2_{i}} + w_{2_{j}}) \cdot \text{Length}_{ij} + & \\
& (\beta_{3} + u_{3_{i}} + w_{3_{j}}) \cdot \text{Exposure}_{ij} + & \\
& (\beta_{4} + u_{4_{i}}) \cdot \text{Cognateness}_{ij} + & \\
& (\beta_{5} + u_{5_{i}} + w_{3_{j}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij}) + & \\
& (\beta_{6} + u_{6_{i}}) \cdot (\text{Age}_{i} \times \text{Cognateness}_{ij}) + & \\
& (\beta_{7} + u_{7_{i}}) \cdot (\text{Exposure}_{ij} \times \text{Cognateness}_{ij}) & \\
& (\beta_{8} + u_{8_{i}}) \cdot (\text{Age}_{i} \times \text{Exposure}_{ij} \times \text{Cognateness}_{ij})
\end{aligned}
$$ {#eq-linear}

where $i$ and $j$ index the participant and translation equivalent (TE), $\beta_{0_k}$ is the fixed coefficient of the regression model for the intercept of threshold $k$, $u_{0_{i}}$ and $w_{0_{j}}$ are the by-participant and by-TE adjustments for $\beta_{0_{k}}$ (i.e., random intercepts), respectively, $\beta_{1-8}$ are the fixed coefficients of the regression model for the predictors of interest, $u_{1-8_{i}}$ and $w_{1-3_{j}}$ are the by-participant and by-TE adjustments for$\beta_{1-8}$ (i.e., random slopes), respectively.

We used the Bayesian framework to estimate the parameters in our model. This involves using the Bayes theorem to compute a distribution (*posterior distribution*) that describes what values of each parameter in the model are more likely given the data (*likelihood*, see @eq-likelihood), and previous knowledge about such distribution (*prior distribution*, see @eq-prior) [@mcelreath2020statistical]. This posterior distribution not only informs about the most likely values of our regression coefficients of interest, but also about the uncertainty around such estimations. We used a weakly informative prior for our parameters, with the exception of the main effect of $\text{Age}$, for which we specified a strongly informative prior based on previous literature about how age affects the acquisition of words (see @eq-prior).

@@

where: $\rho_{u_{0-8}}$ and $\rho_{w_{0-3}}$ indicate the correlation parameters between the by-participant and by-TE adjustments, respectively, $\sigma_{u_{0-8}}^2$ and $\sigma_{w_{0-3}}^2$ indicate the variance of the by-participant and by-TE variance of the adjustments, respectively, and $\mathcal{N}$ indicates a normal distribution, $\mathcal{N}_{+}$ indicates a truncated normal distribution with only positive values, and $LKJcorr$ indicates a [LKJ correlation distribution](https://mc-stan.org/docs/2_22/functions-reference/lkj-correlation.html) [@lewandowski2009generating].

```{r apafg}
#| label: apafg-prior-mean
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 5
#| apa-cap: "Expected prior-predicted mean."
nd <- expand.grid(n_phon_std = 0,
                  age_std = (10:36-mean(responses$age))/(sd(responses$age)),
                  exposure_std = 0,
                  lv_std = 0)

m <- add_epred_draws(nd, 
                     model_fit_4_prior,
                     ndraws = NULL, 
                     re_formula = NA) |> 
    filter(.category != "No") |> 
    pivot_wider(names_from = ".category", values_from = ".epred") |> 
    mutate(Understands = Understands + `Understands and Says`) |> 
    pivot_longer(c(Understands, `Understands and Says`),
                 names_to = ".category",
                 values_to = ".epred") |> 
    mutate(.category = factor(.category,
                              levels = c("Understands and Says",
                                         "Understands"),
                              labels = c("Production\n(Understands and Says)",
                                         "Comprehension\n(Understands)")))

m |> 
    ggplot(aes(x = age_std, y = .epred)) +
    facet_wrap(~.category) +
    stat_lineribbon(size = 0.75, 
                    colour = NA,
                    .width = c(0.95, 0.89, 0.67, 0.5)) +
    stat_summary(fun = "mean", 
                 geom = "line",
                 linewidth = 0.75) +
    scale_x_continuous(breaks = scale(seq(10, 36, 4), 
                                      mean(responses$age), 
                                      sd(responses$age))[, 1],
                       labels = seq(10, 36, 4)) +
    scale_fill_manual(values = rev(clrs), 
                      labels = c("95%", "89%", "67%", "50%")) +
    labs(x = "Age (months)", 
         y = "Probability of acquisition",
         colour = "Credible interval", 
         fill = "Credible interval") +
    scale_y_continuous(labels = percent, limits = c(0, 1)) +
    theme(legend.position = "top",
          axis.ticks.x = element_line(colour = "black"),
          panel.grid = element_blank())
```
